I am infinitely grateful to my supervisors, Dr. Caterina Urban and Alexandra Bugariu, who have provided me with all the support and guidance I needed to complete this project. Thank you for everything you have taught me over the past six months. I could not have hoped for better supervision. \\

I am also very grateful Prof. Dr. Peter M{\"u}ller for the opportunity to work on this project and to be part of a great research environment for half a year. \\

I would also like to thank the lecturers and teaching assistants at the Computer Science Department of the German University in Cairo for everything they have taught me during my studies and for sparking my interest in Computer Science. \\

I am grateful to Gagandeep Singh for helping me incorporate his work on numerical analysis into my work, to Marco Eilers for helping me with a challenging part of my implementation and to J{\'e}r{\^o}me Dohrau for his helpful remarks on my theoretical work. \\

There are no sufficient words with which to express my gratitude to my mother and father whose love and support have shaped my life and whose unwavering faith in me propels me to pursue my dreams. May I always make you proud. \\

I would have never been able to finish this work without the support, encouragement and prayers I recieved from my family and friends in Egypt.\\ 

I am grateful for the frienship of Karin Wenger who showed me around in Switzerland and made my time here unforgettable. \\

Finally, I am grateful to the members of the Chair of Programming Methodoloy at ETH Zurich whose friendly and welcoming attitude made me feel at home as a member of their group for six months. \\

